Endogenous model shifts in algorithmic recourse

### 2022-04-12 dev/students/research-proposal-counterfactual-explanations.qmd
setting up for next steps, including AIES submission

## Project overview

@@

## Motivation and background

@@

## Research questions

@@

Individually, each of the students will then choose one of the recourse generators already implemented in [CARLA](https://github.com/carla-recourse/CARLA) and run a simple experiment (details below) to generate endogenous shifts and quantify their magnitude.
You will also each produce results for the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) and benchmark the results from both generators.

@@

## Methodology

Overall, this will be a project focused on benchmarking and experimental design: you will learn about how to use existing tools for benchmarking to asses outcomes from your own experiments.

1. [FACE](https://arxiv.org/pdf/1909.09369.pdf)
2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf)
3. [CEM](https://arxiv.org/pdf/1802.07623.pdf)
4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf)
5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf)

Since the generators are all conveniently implemented in the same library, running the experiment should require roughly the same effort for all of them. While you are not expected to fully understand in detail the methodology underlying your
chosen generator, you should study the relevant paper and ask yourself: what makes your generator particular? What assumptions do the authors make? How do you expect these particularities to affect recourse outcome with respect to endogenous shifts?

**Experiment**

The idea is that you replicate a variation of the following experiement already implemented by Patrick:

1. Train an algorithm $\mathcal{M}$ for a binary classification task. 
2. Determine a target class. Using your chosen generator and the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) generate recourse for a share $\mu$ of randomly selected individuals in the other class to revise their label (i.e. move to the target class). Store some of the conventional benchmark measures readily implemented in CARLA (cost, success rate, ...)
3. Implement recourse for those indviduals and quantify the domain shift.
4. Retrain $\mathcal{M}$ and quantify the model shift.
5. Repeat steps 1-4 for $K$ rounds.

You are free to either follow Patrick's recipe exactly or tweak the various parameters that determine the experiemntal design as you see fit. Make sure that you end up running the exact same experiment though, in order for the results to be comparable.

**Hypothesis**

- I expect to find that generators that focus on creating realisitic counterfactuals in high-density regions of the feature domain (REVISE, CEM-VAE) tend to be associated with endogenous shifts of a smaller magnitude compared to generators that do not specifically address this (CEM, Wachter, AR-LIME).

## Relation between sub-projects

The students will collectively design and establish an evaluation benchmark, while each focusing on a particular framework implementation to ultimately evaluate in the benchmark. Researching the intricacies of each framework will lead to clearly individual contributions, while part of the relevant literature is shared, and the final comparison connects the individual insights.

## Responsible Engineering / Computer Science / Research

As noted above, algorithmic recourse was designed in the context of responsible machine learning. Work on this topic should foster an awareness that automated decision-making can have detrimental consequences for individuals and that many researchers are working on approaches to deal with some of those consequences. Comparing different approaches to the issue will also make it clear that in the context of research there is rarely (if ever) one right solution. More to the point of this particular research project, solutions to one particular issue usually open up new questions.

## Possible publication

In 2021 there was an [ICML workshop](https://icml.cc/Conferences/2021/ScheduleMultitrack?event=8363) on algorithmic recourse. Should this workshop take place again, the results of this project could lead to a poster presentation. Later in the year NeurIPS is coming up, which in 2021 also attracted conference and workshop papers on algorithmic recourse and benchmarking.

## References

>>>>

### Background and motivation

@@

@@

### Research Questions for the Sub-Projects

@@

Individually, each of the students will then choose one of the recourse generators already implemented in [CARLA](https://github.com/carla-recourse/CARLA) and run a simple experiment (details below) to generate endogenous shifts - that is, provide and implement recourse - and quantify their magnitude. You will also each produce results for the baseline generator ([Wachter](https://arxiv.org/ftp/arxiv/papers/1711/1711.00399.pdf)) and benchmark the results from both generators.

1. [FACE](https://arxiv.org/pdf/1909.09369.pdf), [@poyiadzi2020face]
2. [REVISE](https://arxiv.org/pdf/1907.09615.pdf), [@joshi2019towards]
3. [CEM](https://arxiv.org/pdf/1802.07623.pdf), [@dhurandhar2018explanations]
4. [CEM-VAE](https://arxiv.org/pdf/1802.07623.pdf), [@dhurandhar2018explanations]
5. [AR-LIME](https://arxiv.org/pdf/1809.06514.pdf), [@ustun2019actionable]

@@

### Prerequisites

For this assignment, prior experience is Python is a requirement. A basic understanding of gradient-based optimization is helpful, but not a strict requirement.

### Q&A Session

If you're interested in this project and/or have follow-up questions, please feel free to join us through [Zoom]() on one of these two sessions:

-   Tuesday, March 8, 12:45-13:25 (Zoom [link](https://us04web.zoom.us/j/77787187346?pwd=QFBh8Lc9iliVRzSSa9Yz9XFwZzUrXM.1))
-   Wednesday, March 9, 12:45-13:25 (Zoom [link](https://us04web.zoom.us/j/77845264541?pwd=QoHXseljKn_PMTDawWFrhHPX7AUaua.1))

### References
